Demographic Species Distribution Models

### 2019-10-11 manuscript/ms.Rmd
add more text on alternative specifications

\begin{align}
y_i &\sim Bernoulli(p_i) \\ \nonumber
p_i &= 1 - e^{-N_i} \\ \nonumber
N_i &= \alpha \lambda_i  \\ \nonumber
\lambda_i &= eigval(\mathbf{A_i})
\end{align}

where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the populations at location $i$. Imake a statistically convenient but biologically naive assumption that the size of each population is proportional to its intrinsic growth rate - a relationship that would arise from a constant (not density-dependent) competitive pressure. This simplifying assumption has been made in other spatial population models such as dynamic range models [@pagelForecastingSpeciesRanges2012]. As discussed in the text below, DSDMs can easily incorporate more complex and realistic forms of density-dependence as required.
The parameter $\alpha$ represents the product of this constant of proportionality with the individual-level detection rate (fraction of individuals in the population detected in this year of the survey). This yields $N_i$, the expected number of individuals detected during the survey, which gives the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].

@@

Inference was by Hamiltonian Monte Carlo using the R package greta [@goldingGretaSimpleScalable2019] and its extension package greta.dynamics [@goldingGretaDynamicsModelling2019] for modelling dynamical systems (discussed in *Computation* below). Four chains were run for 1000 iterations each (after 1000 discarded iterations during which the sampler was burnt-in and tuned), which was sufficent to ensure convergence (`r sprintf("$\\hat{R}\\le%s$", format(bbs_max_rhat, digits = 3))`), and yield at least `r round(bbs_neff)` effective samples, for all model parameters. This took `r round(bbs_time)` seconds on a laptop computer. Table 1 reports estamited posterior distributions over these parameters.

@@

The fitted DSDM yields predictions of the probability of presence like those from a correlative SDM (Fig. 2A): predictions from a maximum likelihood logistic regression model fitted to the same data had a correlation of `r bbs_stats$glm_dsdm_correl` with the DSDM predictions, and similar AUC and correlation when compared with the true data in the `r bbs_n_test` survey locations not used to train the model (GLM: AUC = `r bbs_stats$glm_auc`, correlation = `r bbs_stats$glm_pa_correl`, DSDM: AUC = `r bbs_stats$dsdm_auc`, correlation = `r bbs_stats$dsdm_pa_correl`).
Unlike a correlative SDM, the DSDM also predicts spatial variation in the survival (Fig. 2B) and fecundity (Fig. 2C) rates.
The species’ range limits (in this case due to the fundamental niche) can be mapped by identifying areas where $\lambda_i \ge 1$, and the model can predict the distribution under reduced fecundity (e.g. due to an introduced competitor) or survival (e.g. due to an introduced pathogen) (Fig. 2D).

@@

Note that this model could undoubtedly be improved by including more information of the species ecology and vital rates - for example constraining spatial estimates of fecundity and survival to within a range of plausible values - via prior distributions or changes to model structure. As with any model, the reliability of these vital rate and distribution predictions will depend on the assumptions made and information provided by the modeller. Unlike in purely correlative models however, these assumptions are made clear by explicitly considering spatiall-yvarying vital rates.

# Extending the basic DSDM

Basic DSDMs like the model in Example 1 do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
Where such information is available (e.g. from studies or expert opinion) it can be incorporated to improve DSDM predictions either via informative priors or by directly incorporating predictions from other models.
The DSDMs I propose are based on transition matrices (including the matrices used in both MPMs and IPMs), which are well studied in ecology and have been used to study a wide range of plant and animal species with very different life histories [@ellnerIntegralProjectionModels].
Like other transition matrix approaches, the basic DSDMs I have already developed can also be extended to explicitly model three key ecological processes that are poorly captured by existing distribution models: density dependence, dispersal, and biotic interactions.

@@

@@

@@

@@

>>>>

\begin{align}
y_i &\sim Bernoulli(p_i) \\ \nonumber
p_i &= 1 - e^{-\Lambda_i} \\ \nonumber
\Lambda_i &= \alpha \lambda_i  \\ \nonumber
\lambda_i &= eigval(\mathbf{A_i})
\end{align}

where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the populations at location $i$. Imake a statistically convenient but biologically naive assumption that the size of each population is proportional to its intrinsic growth rate - a relationship that would arise from a constant (not density-dependent) competitive pressure. This simplifying assumption has been made in other spatial population models such as dynamic range models [@pagelForecastingSpeciesRanges2012]. As discussed in the text below, DSDMs can easily incorporate more complex and realistic forms of density-dependence as required.
The parameter $\alpha$ represents the product of this constant of proportionality with the individual-level detection rate (fraction of individuals in the population detected in this year of the survey). This yields $\Lambda_i$, the expected number of individuals detected during the survey, which gives the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].

@@

Inference was by Hamiltonian Monte Carlo using the R package greta [@goldingGretaSimpleScalable2019] and its extension package greta.dynamics [@goldingGretaDynamicsModelling2019] for modelling dynamical systems (discussed in *Computation* below). Four chains were run for 1000 iterations each (after 1000 discarded iterations during which the sampler was burnt-in and tuned), which was sufficent to ensure convergence (`r sprintf("$\\hat{R}\\le%s$", format(bbs_max_rhat, digits = 3))`), and yield at least `r round(bbs_neff)` effective samples, for all model parameters. This took `r round(bbs_time)` seconds on a laptop computer. Table 1 reports estimates of the posterior distributions over these parameters.

@@

Prediction of the DSM to new locations proceeds as with any Bayesian model by substituting the design matrix $\mathbf{X}$ for observed data with a design matrix corresponding to the prediction locations, and for each posterior sample of parameter values, calculating the value of the variable of interest. This yields a posterior distribution over predictions that incorporates uncertainty in parameter values. 
The fitted DSDM yields (posterior mean) predictions of the probability of presence similar to those from a correlative SDM (Fig. 2A): predictions from a maximum likelihood logistic regression model fitted to the same data had a correlation of `r bbs_stats$glm_dsdm_correl` with the DSDM predictions, and similar AUC and correlation when compared with the true data in the `r bbs_n_test` survey locations not used to train the model (GLM: AUC = `r bbs_stats$glm_auc`, correlation = `r bbs_stats$glm_pa_correl`, DSDM: AUC = `r bbs_stats$dsdm_auc`, correlation = `r bbs_stats$dsdm_pa_correl`).
Unlike a correlative SDM, the DSDM also predicts spatial variation in the survival (Fig. 2B) and fecundity (Fig. 2C) rates.
The species’ range limits (in this case due to the fundamental niche) can be mapped by identifying areas where $\lambda_i \ge 1$, and the model can predict the distribution under reduced fecundity (e.g. due to an introduced competitor) or survival (e.g. due to an introduced pathogen) (Fig. 2D).

@@

Note that this model could undoubtedly be improved by including more information of the species ecology and vital rates - for example constraining spatial estimates of fecundity and survival to within a range of plausible values - via prior distributions or changes to model structure. As with any model, the reliability of these vital rate and distribution predictions will depend on the assumptions made and information provided by the modeller. Unlike in purely correlative models however, these assumptions are made clear by explicitly considering spatially-varying vital rates.

# Alternative DSDM specifications

The DSDM structure given in equations 2-3 is the simplest version. It can easily be extend to accommodate alternative data types and demographic processes, and answer a range of ecological questions.

## Alternative observation models

DSDMs explicitly consider the density of individuals at each location through the relative abundance variable $\Lambda_i$. This provides a link to the most commonly used observations for modelling species distributions in ecology.
An observation model for count data can be specified simply by replacing the bernoulli sampling distribution in equation 3 with an appropriatecount (such as a Poisson or negative binomial distribution) with $\Lambda_i$ as the expected count.
A model for presence-only data can be formed by considering the data as arising from a Poisson point process with $\Lambda_i$ as the intensity at location $i$ [@rennerPointProcessModels2015]. Integration areas for each observation, and any modelled or prespecified sampling bias can be provided via $\alpha$.
Similarly, where repeat surveys are available, a model for $\alpha$ can be used to estimate the detection probability as in an occupancy-detection analysis [@mackenzieEstimatingSiteOccupancy2002].

# Temporal dynamics

The example DSDM given above is temporally static: the covariate values, vital rate parameters, transition matrix and intrinsic growth rate at each site are assumed to be constant through time.
Long-term temporal trends driven by changing environmental covariates can be accommodated in this model simply be taking the index $i$ to represent observations at distinct locations *and times*, with the observation vector and design matrix extended accordingly.
Shorter-term temporal trends and the impacts of environmental stochasticity on species persistence can also be included by using of dynamic (temporally-varying) transition matrices:

\begin{align}
\Lambda_{t,i} &= \alpha N_{t,i}  \\ \nonumber
N_{t,i} &= \mathbf{A(\mathbf{X}_{t,i})}N_{t-1,i}
\end{align}

where the new variable $N_{t,i}$ gives the expected population size at time $t$, which is also used as an index to environmental covariates $\mathbf{X}$ and observations. Rather than the intrisic growth rate computed form the static matrix, this formulation explicitly calculates population size at each time $t$ from the deterministic dynamics that define the matrix population model. In this formulation of the DSDM, the initial population at each site $N_{0,i}$ must be provided. This could either be specified as a parameter in the model, via a known initial population size, or as an arbitrary fixed value accompanied by a sufficient 'burn in' period that parameter estimates are not influenced.

# Continuous stages

Can use IPMs too...

The DSDMs I propose are based on transition matrices (including the matrices used in both MPMs and IPMs), which are well studied in ecology and have been used to study a wide range of plant and animal species with very different life histories [@ellnerIntegralProjectionModels].

@@

 add notation for A(N_t)
 note that this can be combined with Eq 5, or can be iterated for a sufficient number of timesteps (T) to get a temporally-static density-dependent model. In practice, T can be determined adaptively, iterating a sufficient number of times for N_i to converge and then terminating.

@@

Two approaches (with equation)
- for models using the eigenvalue convolve A with a dispersal matrix D (using a mask for dispersing stages)
- or for density-dependent or temporally dynamic models interleave dispersal and reproduction stages
note that in practice we would compute the eigenvalue by iteration for computational convenience.

Note that the initial populations can be specified in only some patches to represent dispersal limitation. For example in the CaCh example setting zero initial population in the pacific north west - where there is a patch of apparently suitable habitat but where the species does not exist (presumably due to dispersal limitation).

@@

add example matrix structure as an equation

@@

# Prior specification

Basic DSDMs like the model in Example 1 do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
Where such information is available (e.g. from studies or expert opinion) it can be incorporated to improve DSDM predictions either via informative priors or by directly incorporating predictions from other models.

Determining the most common and useful types of prior information, and coming up with straightforward strategies to incorporate them is an area for future research.

The affine transformation provided in example 1 is one approach. An integrated likelihood is another option - the analysis of recapture data to inform vital rates at specific sites could be performed simultaneously with estimation of the DSDM...
